Airspace configuration decision based on Monte-Carlo tree search

### 2018-04-20 article/article.tex
(-) no more custom commands
(+) subsubsub
(~) Differentiated notations
Algorithms systematically use nodes whereas equations use states.

\DeclareMathOperator{\argmax}{argmax}
\DeclareMathOperator{\cons}{::}

@@

@@

\section*{Notations}
\begin{itemize}
  \item $m$: elementary module
  \item $S$: sector control position, group of modules
  \item $P$: partition of the airspace, group of sectors
  \item $t$: current time
  \item $C(P, t)$: cost of a given partition $P$ at time $t$
  \item $s$: state, in this article a partition $P$ and a time $t$
  \item $h(s)$: heuristic that gives for a state $s$ an estimation of the cost
  \item $\mu_{v, a}$: mean of all reward comming from node $v$ with action $a$
  \item $T_v$: number of simulation where the final path include the node $v$
  \item $T_{v, a}$: number of simulation where the final path include the
    node $F$ with action $a$
  \item \(\mathcal{A}(v)\) the available actions from a node \(v\)
  \item \(p\colon \mathcal{S} \times \mathcal{A} \to \mathcal{S}\) the
    production function, giving the resulting state of action \(a\) applied on a
    state \(s\).
\end{itemize}

@@

\subsection{Monte Carlo tree search}
\subsubsection{Overview}
Monte-Carlo is a best-first search method using stochastic simulations. The
algorithm actually uses two trees, an underlying tree associated to the model
(e.g.\ a game tree) and a search tree. The latter is built incrementally by the
algorithm the following way, at each step, starting from the root, the most
promising nodes towards the bottom of the tree are selected. Once a leaf of the
search tree is reached, a new node coming from the model tree is added to the
leaf. A simulation is then run from this newly added node to evaluate the
outcome of a path starting from it. The result of the simulation is eventually
backpropagated to the consecutive fathers of the new node.

The above mentioned steps are denoted
\begin{compactenum}
\item selection (or tree walk): choosing successively most promising nodes,
\item expansion: adding new nodes to the search tree,
\item simulation (or random walk): choosing quickly actions from the model
  tree,
\item backpropagation (or backup): applying the result of the simulation to
  the search tree.
\end{compactenum}
Those phases are repeated until a stopping criterium (e.g.\ memory or time) is
reached, resulting in algorithm~\ref{alg:gen_mcts}.
\begin{algorithm}
  \caption{General MCTS~\cite{browne2012survey}}\label{alg:gen_mcts}
  \begin{algorithmic}
    \Procedure{MctsSearchTree}{$v_0$}
    \While{within computational budget}
    \State{} \([v_0, \dots, v_n] \gets\) \Call{TreePolicy}{$v_0$}
    \State{} \(\Delta \gets\) \Call{DefaultPolicy}{$v_n$}
    \State{} \Call{Backup}{$[v_0, \dots, v_n]$, $\Delta$}
    \EndWhile{}
    \EndProcedure{}
  \end{algorithmic}
\end{algorithm}

\subsubsection{Extending the search tree}
\paragraph{Selection~\ref{alg:tree_pol}}
The aim of the selection is to choose from a set of actions, the most promising
one\footnote{i.e.\ possibly leading to the best evaluation} according to a
\emph{tree policy}. This type of problem is addressed by bandits methods where
a multi armed bandit aims to always activate the slot machine giving the highest
reward.

The bandit problem is applied to MCTS via the Upper Confidence Tree algorithm.
It uses the UCB for Upper Confidence Bound as tree policy (introduced
in~\cite{kocsis2006bandit}). Let \(v\) be the node from which an action must be
selected to go deeper in the tree, \(a\) an action available from \(v\),
\(\mu_{v,a}\) the expected outcome from the node obtained by applying action
\(a\) on \(v\), \(T_v\) the number of simulations carried out from node \(v\)
(which includes any simulation from nodes in any subtree of \(v\)) and
\(T_{v,a}\) the number of simulations carried out from the node resulting from
the action \(a\) applied to node \(v\), \(\beta\) a constant. The selected node
is thus the one maximising
\begin{equation}
  \text{UCB1} = \mu_{v,a} + 2 \beta \sqrt{\frac{2 \log T_v}{T_{v,a}}}
\end{equation}

While the previous equation is well suited for two players games, it can be
tweaked to improve its efficience in one player games or sequencing problems. An
alternative using the standard deviation is proposed in~\cite{sebag2010fuse}
called the \emph{UCB1-tuned} equation
\begin{equation}
  \mu_{v,a} +
  \sqrt{%
    \frac{\beta\log T_v}{t_{v,a}}
    \min\left( \frac{1}{4}, \sigma^2_{v,a} +
    \sqrt{%
      \frac{2 \log T_v}{T_{v,a}}
    }\right)
  }
\end{equation}
\begin{algorithm}
  \caption{UCT algorithm}\label{alg:tree_pol}
  \begin{algorithmic}
    \Function{TreePolicy}{path of nodes $\pi$}
    \State{} $v \gets$ head of $\pi$
    \If{$v$ is terminal}
    \State{}\Return{$\pi$}
    \Else{}
    \If{$v$ is not fully expanded}
    \State{}\Return{\Call{Expand}{$v$} $\cup \pi$}
    \Else{}
    \State{}$f \gets$ \Call{BestChild}{$v$}
    \State{}\Return{%
      \Call{TreePolicy}{$f \cup \pi$}
    }
    \EndIf{}
    \EndIf{}
    \EndFunction{}
  \end{algorithmic}
  \begin{algorithmic}
    \Function{BestChild}{$v$}
    \State{}\Return{$\argmax\{\Call{UCB}{v'} |
    v' \text{\,children of\,} v\}$}
    \EndFunction{}
  \end{algorithmic}
\end{algorithm}

\paragraph{Expansion~\ref{alg:expansion}}
The above step is carried out only if, considering a node \(v\), a simulation
has been run through each child available from \(v\). In the case where one or
more children haven't been explored yet, one of the unexplored children is
chosen and a simulation is started from it.
\begin{algorithm}
  \caption{%
    Expansion (\(p\colon \mathcal{S} \times \mathcal{A} \to \mathcal{S}\)
    is the transition function)
  }\label{alg:expansion}
  \begin{algorithmic}
    \Function{Expand}{node $v$}
    \State{}$\ell \gets \{p(v, a) | a\in\mathcal{A}(v), T_{v,a} = 0\}$
    \State{}\Return{random element from $\ell$}
    \EndFunction{}
  \end{algorithmic}
\end{algorithm}

@@

\paragraph{Backpropagation}
The backpropagation consists in updating the values required to carry out the
tree policy. The values must be updated incrementally since the backpropagation
function has the current value of the parameters and the result of the
simulation as parameters. Thus, for UCB1 equation, the expected reward (mean)
and the total count are updated this way
\begin{algorithm}
  \caption{UCB1 backpropagation}
  \begin{algorithmic}
    \Procedure{Backpropagate}{$v$, $r$}
    \State{}\(\delta \gets\ r - \mu_v\)
    \State{}\(\mu_v \gets \mu_v + \frac{\delta}{T_v + 1}\)
    \State{}\(T_v \gets T_v + 1\)
    \EndProcedure{}
  \end{algorithmic}
\end{algorithm}

For the UCB1-tuned, the standard deviation has to be computed. It results in
algorithm~\ref{alg:backpropagate-tuned} which introduces the value \(m_v\)
associated to a node \(v\) to compute the standard deviation via the formula
\begin{equation}
  \sigma^2_v = \frac{m_v}{T_v}.
\end{equation}
\begin{algorithm}
  \caption{UCB1-tuned backpropagation}\label{alg:backpropagate-tuned}
  \begin{algorithmic}
    \Procedure{Backpropagate}{$v, r$}
    \State{}\(\delta \gets r - \mu_v\)
    \State{}\(\mu_v \gets \mu_v + \frac{\delta}{T_v + 1}\)
    \State{}\(\delta' \gets r - \mu_v\)
    \State{}\(m_v \gets m_v + \delta \delta'\)
    \State{}\(T_v \gets T_v + 1\)
    \EndProcedure{}
  \end{algorithmic}
\end{algorithm}

\subsection{Sequence building}
\subsubsection{Node selection}\label{sssec:node_selection}
The building of the final sequence of actions differs from the selection on the
aim, the selection chooses the most promising actions while the final sequence
must be composed of the best. This means that the final path build won't use the
UCB equation. Chaslot \textit{et al.}~\cite{browne2012survey} propose several
methods to select nodes,
\begin{itemize}
  \item max-child: select the child with highest reward;
  \item robust child: select the most visited root child;
  \item  secure child: select the child maximising a lower confidence bound.
\end{itemize}

\subsubsection{Iterative pathfinding}
The path is built iteratively by calling successive Monte Carlo tree searches.
Say an MCTS has been called on a node \(u\). Once a stopping criterium is
matched, a node \(v\) among the ones reachable from \(u\) is selected via one of
the previously mentioned policies. Then an mcts is called back with node \(u\)
as the root node. The path is composed of the successive roots. The algorithm is
described at~\ref{alg:path_building}.
\begin{algorithm}
  \caption{%
    Path building. The {\sc NextNode} (here max-child) function is one among
    those in~\ref{sssec:node_selection}.
  }\label{alg:path_building}
  \begin{algorithmic}
    \Function{MctsSearch}{$u_0, n$}
    \State{}\(u \gets u_0\)
    \State{}\(\pi \gets \{u\}\)
    \For{$i = 1$ to $n$}
      \State{}\Call{MctsSearchTree}{$u$}
      \State{}\(u \gets\)\Call{NextNode}{$u$}
      \State{}\(\pi \gets \pi \cup \{u\}\)
    \EndFor{}
    \State{}\Return{$\pi$}
    \EndFunction{}
    \Function{NextNode}{$u$}
    \State{}\Return{$\argmax\{\mu_{u,a} | a \in \mathcal{A}(v)\}$}
    \EndFunction{}
  \end{algorithmic}
\end{algorithm}
\subsection{A\(^*\)}
To evaluate the exactness of the paths given by the MCTS algoithm, the A\(^*\)
algorithm is used. A\(^*\) is an exact depth first search pathfinding algorithm
which uses a heuristic function to guide its search. The algorithm is given
in~\ref{alg:astar} where \(u_0\) is the initial state, \(T\) the set of termial
nodes, \(h\) a heuristic function estimating the cost from a state \(u\) given
as argument to a final state, \(k \colon \mathcal{S}^2 \to \mathbb{R}\). The
function \(\mathup{first(G)}\) returns the first element of \(G\) and
\(f\mathup{-insert}(G, v)\) inserts \(v\) in \(G\) ordering first by \(f\)
increasing then by \(g\) decreasing.
\begin{algorithm}
  \caption{A\(^*\) algorithm~\cite{alliotschiex2002ia&it}}\label{alg:astar}
  \begin{algorithmic}
    \Procedure{A$^*$}{$u_0$}
    \State{}\(G \gets u_0; D \gets \emptyset; g(u_0) \gets 0; f(u_0) \gets 0\)
    \State{}\(\mathup{father}(u_0) \gets \emptyset\)
    \While{$G \neq \emptyset$}
    \State{}\(u \gets \mathup{first}(G); G \gets G \backslash \{u\}\)
    \State{}\(D \gets D \cup \{u\}\)
    \If{$u \in T$}
    \State{}\Return{father}
    \EndIf{}
    \For{$v$ in childen of $u$}
    \If{$v \notin D \cup G$ or $[g(v) > g(u) + k(u, v)]$}
    \State{}\(g(v) \gets g(u) + k(u, v)\)
    \State{}\(f(v) \gets g(v) + h(v)\)
    \State{}\(\mathup{father}(v) \gets u\)
    \State{}\(f\mathup{-insert}(G, v)\)
    \EndIf{}
    \EndFor{}
    \EndWhile{}
    \EndProcedure{}
  \end{algorithmic}
\end{algorithm}
\section{Model}

>>>>

\DeclareMathOperator{\argmax}{argmax}
\DeclareMathOperator{\first}{first}
\DeclareMathOperator{\state}{state}

@@

\section*{Notations}
\begin{itemize}
  \item $m$: elementary module;
  \item $S$: sector control position, group of modules;
  \item $P$: partition of the airspace, group of sectors;
  \item $t$: current time;
  \item $C(P, t)$: cost of a given partition $P$ at time $t$;
  \item \(\mathcal{S}\): the set of available states;
  \item \(\mathcal{A}\): the set of all possible actions;
  \item \(\mathcal{A}(s)\): the available actions from a state \(s\);
  \item $s$: state;
  \item \(f \colon \mathcal{S} \times \mathcal{A} \to \mathcal{S}\):
    transition function giving the resulting states of action \(a\) executed
    from a state \(s\).
  \item $h \colon \mathcal{S} \to \mathbb{R}^+$: heuristic estimating the cost
    from a state \(s \in \mathcal{S}\) to a final state;
  \item \(\mathcal{N}\): the set of nodes
  \item \(u, v \in \mathcal{N}^2\): nodes
  \item \(Q \colon \mathcal{N} \to \mathbb{R}^+\): the reward expectancy of a
    node;
  \item \(N \colon \mathcal{N} \to \mathbb{N}\): number of simulations running
    through or from a node;
  \item \(\first(\ell)\): returns the first element of \(\ell\).
\end{itemize}

@@

@@

\subsection{Monte Carlo tree search}
\subsubsection{Overview}
Monte-Carlo is a best-first search method using stochastic simulations. The
algorithm actually uses two trees, an underlying tree associated to the model
(e.g.\ a game tree) and a search tree. The latter is built incrementally by the
algorithm the following way, at each step, starting from the root, the most
promising nodes towards the bottom of the tree are selected. Once a leaf of the
search tree is reached, a new node coming from the model tree is added to the
leaf. A simulation is then run from this newly added node to evaluate the
outcome of a path starting from it. The result of the simulation is eventually
backpropagated to the consecutive parents of the new node.

The above mentioned steps are denoted
\begin{compactenum}
\item selection (or tree walk): choosing successively most promising nodes,
\item expansion: adding new nodes to the search tree,
\item simulation (or random walk): choosing quickly actions from the model
  tree,
\item backpropagation (or backup): applying the result of the simulation to
  the search tree.
\end{compactenum}
Those phases are repeated until a stopping criterium (e.g.\ memory or time) is
reached, resulting in algorithm~\ref{alg:gen_mcts} where the tree policy
aggregates phases 1 and 2 to create a new node of the search tree and the
default policy gives an evaluation of the newly added node.
\begin{algorithm}
  \caption{General MCTS~\cite{browne2012survey}}\label{alg:gen_mcts}
  \begin{algorithmic}
    \Procedure{MctsSearchTree}{$v_0$}
    \While{within computational budget}
    \State{} \([v_0, \dots, v_n] \gets\) \Call{TreePolicy}{$v_0$}
    \State{} \(\Delta \gets\) \Call{DefaultPolicy}{$v_n$}
    \State{} \Call{Backup}{$[v_0, \dots, v_n]$, $\Delta$}
    \EndWhile{}
    \EndProcedure{}
  \end{algorithmic}
\end{algorithm}

\subsubsection{Extending the search tree}
\paragraph{Selection (algorithm~\ref{alg:tree_pol})}
The aim of the selection is to choose from a set of actions, the most promising
one\footnote{i.e.\ possibly leading to the best evaluation} according to a
\emph{tree policy}. This type of problem is addressed by bandits methods where
a multi armed bandit aims to always activate the slot machine giving the highest
reward.

The bandit problem is applied to MCTS via the Upper Confidence Tree algorithm.
It uses the UCB (Upper Confidence Bound) as tree policy (introduced
in~\cite{kocsis2006bandit}). Let \(v\) be the node from which an action must be
selected to go deeper in the tree, \(a\) an action available from \(v\),
\(\mu_{v,a}\) the expected outcome from the node obtained by applying action
\(a\) on \(v\), \(T_v\) the number of simulations carried out from node \(v\)
(which includes any simulation from nodes in any subtree of \(v\)) and
\(T_{v,a}\) the number of simulations carried out from the node resulting from
the action \(a\) applied to node \(v\), \(\beta\) a constant. The selected node
is thus the one maximising the \emph{UCB1} equation
\begin{equation}
  \mu_{v,a} + 2 \beta \sqrt{\frac{2 \log T_v}{T_{v,a}}}
\end{equation}

While the previous equation is well suited for two players games, it can be
tweaked to improve its efficience in one player games or sequencing problems. An
alternative using the standard deviation of the outcome of the simulations is
proposed in~\cite{sebag2010fuse} called the \emph{UCB1-tuned} equation
\begin{equation}
  \mu_{v,a} +
  \sqrt{%
    \frac{\beta\log T_v}{t_{v,a}}
    \min\left( \frac{1}{4}, \sigma^2_{v,a} +
    \sqrt{%
      \frac{2 \log T_v}{T_{v,a}}
    }\right)
  }
\end{equation}
\begin{algorithm}
  \caption{UCT algorithm}\label{alg:tree_pol}
  \begin{algorithmic}
    \Function{TreePolicy}{$\pi$}
    \State{}\(v \gets \first(\pi)\)
    \If{$v$ is terminal}
    \State{}\Return{$\pi$}
    \Else{}
    \If{$v$ is not fully expanded}
    \State{}\Return{\Call{Expand}{$v$} $\cup \pi$}
    \Else{}
    \State{}$f \gets$ \Call{BestChild}{$v$}
    \State{}\Return{%
      \Call{TreePolicy}{$f \cup \pi$}
    }
    \EndIf{}
    \EndIf{}
    \EndFunction{}
  \end{algorithmic}
  \begin{algorithmic}
    \Function{BestChild}{$v$}
    \State{}\Return{$\argmax\{\Call{UCB}{v'} | v' \text{\,children of\,} v \}$}
    \EndFunction{}
  \end{algorithmic}
\end{algorithm}

\paragraph{Expansion (algorithm~\ref{alg:expansion})}
The above step is carried out only if, considering a node \(v\), a simulation
has been run through each child available from \(v\). In the case where one or
more children haven't been explored yet, one of the unexplored children is
chosen and a simulation is started from it.
\begin{algorithm}
  \caption{Expansion}\label{alg:expansion}
  \begin{algorithmic}
    \Function{Expand}{$v$}
    \State{}$\ell \gets \{ v | v \text{\,children of\,} u, N(v) = 0 \}$
    \State{}\Return{random element from $\ell$}
    \EndFunction{}
  \end{algorithmic}
\end{algorithm}

@@

\paragraph{Backpropagation}
The backpropagation consists in updating the values required to carry out the
tree policy. The values must be updated incrementally since the backpropagation
function has the current value of the parameters and the result of the
simulation as parameters. Thus, for UCB1 equation, the expected reward (mean)
and the total count are updated this way
\begin{algorithm}
  \caption{UCB1 backpropagation}
  \begin{algorithmic}
    \Procedure{Backpropagate}{$v$, $r$}
    \State{}\(\delta \gets\ r - \mu_v\)
    \State{}\(Q(v) \gets Q(v) + \frac{\delta}{N(v) + 1}\)
    \State{}\(N(v) \gets N(v) + 1\)
    \EndProcedure{}
  \end{algorithmic}
\end{algorithm}

For the UCB1-tuned, the standard deviation has to be computed. It results in
algorithm~\ref{alg:backpropagate-tuned} which introduces the value \(m_2(v)\)
associated to a node \(v\) to compute the standard deviation via the formula
\begin{equation}
  \sigma^2_v = \frac{m_2(v)}{N(v)}.
\end{equation}
\begin{algorithm}
  \caption{UCB1-tuned backpropagation}\label{alg:backpropagate-tuned}
  \begin{algorithmic}
    \Procedure{Backpropagate}{$v, r$}
    \State{}\(\delta \gets r - Q(v)\)
    \State{}\(Q(v) \gets Q(v) + \frac{\delta}{N(v) + 1}\)
    \State{}\(\delta' \gets r - Q(v)\)
    \State{}\(m_2(v) \gets m_2(v) + \delta \delta'\)
    \State{}\(N(v) \gets N(v) + 1\)
    \EndProcedure{}
  \end{algorithmic}
\end{algorithm}

\subsection{Sequence building}
\subsubsection{Node selection}\label{sssec:node_selection}
The building of the final sequence of actions differs from the selection on the
aim, the selection chooses the most promising actions while the final sequence
must be composed of the best. This means that the final path build won't use the
UCB equation. Chaslot \textit{et al.}~\cite{browne2012survey} propose several
methods to select nodes,
\begin{compactitem}
  \item max-child: select the child with highest reward;
  \item robust child: select the most visited root child;
  \item secure child: select the child maximising a lower confidence bound.
\end{compactitem}

\subsubsection{Iterative pathfinding}
The path is built iteratively by calling successive Monte Carlo tree searches.
Say an MCTS has been called on a node \(u\). Once a stopping criterium is
matched, a node \(v\) among the ones reachable from \(u\) is selected via one of
the previously mentioned policies. Then an mcts is called back with node \(u\)
as the root node. The path is composed of the successive roots. The algorithm is
described at~\ref{alg:path_building}.
\begin{algorithm}
  \caption{%
    Path building. The {\sc NextNode} (here max-child) function is one among
    those in~\ref{sssec:node_selection}.
  }\label{alg:path_building}
  \begin{algorithmic}
    \Function{MctsSearch}{$u_0, n$}
    \State{}\(u \gets u_0\)
    \State{}\(\pi \gets \{u\}\)
    \For{$i = 1$ to $n$}
      \State{}\Call{MctsSearchTree}{$u$}
      \State{}\(u \gets\)\Call{NextNode}{$u$}
      \State{}\(\pi \gets \pi \cup \{u\}\)
    \EndFor{}
    \State{}\Return{$\pi$}
    \EndFunction{}
    \Function{NextNode}{$u$}
    \State{}\Return{$\argmax\{Q(v) | v \text{\,children of\,} u\}$}
    \EndFunction{}
  \end{algorithmic}
\end{algorithm}
\subsection{A\(^*\)}
To evaluate the exactness of the paths given by the MCTS algoithm, the A\(^*\)
algorithm is used. A\(^*\) is an exact depth first search pathfinding algorithm
which uses a heuristic function to guide its search. The algorithm is given
in~\ref{alg:astar} where \(u_0\) is the initial state, \(T\) the set of termial
nodes, \(h\) a heuristic function estimating the cost from a state \(u\) given
as argument to a final state, \(k \colon \mathcal{S}^2 \to \mathbb{R}\). The
function \(\mathup{first(G)}\) returns the first element of \(G\) and
\(f\mathup{-insert}(G, v)\) inserts \(v\) in \(G\) ordering first by \(f\)
increasing then by \(g\) decreasing.
\begin{algorithm}
  \caption{A\(^*\) algorithm~\cite{alliotschiex2002ia&it}}\label{alg:astar}
  \begin{algorithmic}
    \Procedure{A$^*$}{$u_0$}
    \State{}\(G \gets u_0; D \gets \emptyset; g(u_0) \gets 0; f(u_0) \gets 0\)
    \State{}\(\mathup{father}(u_0) \gets \emptyset\)
    \While{$G \neq \emptyset$}
    \State{}\(u \gets \mathup{first}(G); G \gets G \backslash \{u\}\)
    \State{}\(D \gets D \cup \{u\}\)
    \If{$u \in T$}
    \State{}\Return{father}
    \EndIf{}
    \For{$v$ in childen of $u$}
    \If{$v \notin D \cup G$ or $[g(v) > g(u) + k(u, v)]$}
    \State{}\(g(v) \gets g(u) + k(u, v)\)
    \State{}\(f(v) \gets g(v) + h(v)\)
    \State{}\(\mathup{father}(v) \gets u\)
    \State{}\(f\mathup{-insert}(G, v)\)
    \EndIf{}
    \EndFor{}
    \EndWhile{}
    \EndProcedure{}
  \end{algorithmic}
\end{algorithm}
\section{Model}
